---
title: "RMS titanic analysis"
author: "chinmay gupta"
date: "16 January 2018"
output: html_document
---

# task 2(a)
Download and review the Titanic Data.csv data file associated with some Titanic Survivors.

#task 2(b)
Use the read.csv() function in R to read the data and store it in a dataframe called "titanic".

Use the View() function in R to view the dataframe in R
```{r}
titanic.df <- read.csv(paste("Titanic Data.csv"), sep= ",")
View(titanic.df)
```

#task 3(a)
 count the total number of passengers on board the Titanic.
 
```{r}
nrow(titanic.df)
```

#task 3(b)
count the number of passengers who survived the sinking of the Titanic.

```{r}
nrow(subset(titanic.df,Survived==1))
```

#task 3(c)
measure the percentage of passengers who survived the sinking of the Titanic.

```{r}
(prop.table(table(titanic.df$Survived))*100)[2]
```

# task 3(d)

count the number of first-class passengers who survived the sinking of the Titanic.

```{r}
mytable <- xtabs(~Survived+Pclass,data=titanic.df)
mytable[2]
```

#task 3(e)
measure the percentage of first-class passengers who survived the sinking of the Titanic.

```{r}
(prop.table(mytable)*100)[2]
```

#task 3(f)
count the number of females from First-Class who survived the sinking of the Titanic
```{r}
female <- xtabs(~Survived+Pclass+Sex,data=titanic.df)
(ftable(female))[4]
```

#task 3(g)
measure the percentage of survivors who were female

```{r}
mytable <- xtabs(~Survived+Sex,data=titanic.df)
(prop.table(mytable,1)*100)[2,1]

```

#task 3(h)
measure the percentage of females on board the Titanic who survived.

```{r}
(prop.table(mytable,2)*100)[2,1]
```


#task 3(i)
Run a Pearson's Chi-squared test to test the following hypothesis:

Hypothesis:  The proportion of females onboard who survived the sinking of the Titanic was higher than the proportion of males onboard who survived the sinking of the Titanic.


```{r}
chisq.test(mytable)
```



since the P-value < 0.05, we reject the null hypothesis.















 
