---
title: "Titanic Survival Analysis"
author: "Sameer Mathur"
output:
  pdf_document: default
  html_document: default
---


Read the Titanic dataset
========================================================

```{r}
titanic <- read.csv(paste("Titanic Data.csv", sep=""))
head(titanic)  # first few rows of the data frame
```

Random selection of rows using car::some() function
========================================================

```{r}
library(car)   
some(titanic)  # random rows from the data frame
```

Structure of the data frame
========================================================

```{r}
str(titanic)   # datatypes of the columns in data frame
```

Summary Statistics of the data
========================================================

```{r, message=FALSE, warning=FALSE}
attach(titanic)   # attaching the columns of the data frame
library(psych)
describe(titanic)   # using describe function
```

Dimension of the data frame
========================================================

```{r}
dim(titanic)   # Total number of rows and columns in data frame
```


QUESTION 1
========================================================
Q1a. How many passengers were on board the Titanic? (Hint: The R function `dim()` may be helpful)

Q1b. How many passengers survived the sinking of the Titanic? (Hint: The R functions <span style="color:blue">`subset()`</span> or <span style="color:blue">`which()`</span> may be helpful)

Q1c. Create a one-way contingency table summarizing the Titanic passengers based on how many survived and how many died. (Hint: The R functions <span style="color:blue">`as.factor()`</span> and <span style="color:blue">`table()`</span> may be helpful)

Q1d. What was the percentage of passengers who survived the sinking of the Titanic? (Hint: The R function <span style="color:blue">`prop.table()`</span> may be helpful)


Q1a. How many passengers were on board the Titanic?
========================================================

```{r}
# total number of passengers
dim(titanic)[1]
```

Q1b. How many passengers survived the sinking of the Titanic?
========================================================

```{r}
# using subset() function 
nrow(subset(titanic, Survived == 1))

# using which() function
length(which(Survived == 1))
```

Q1c. Create a one-way contingency table of the Titanic passengers based on who survived and who died.
========================================================

```{r}
titanic$Survived <- as.factor(titanic$Survived)
# number of passengers who survived and who died
survivedTable <- table(titanic$Survived)
survivedTable
```

Q1d. What was the percentage of passengers who survived the sinking of the Titanic?
========================================================

```{r}
# proportion of survived and died
prop <- prop.table(survivedTable)

# percentage
propPer <- prop*100
propPer
# percentage of passengers who survived 
propPer[2]
```


QUESTION 2
========================================================

Q2a. Create a two-way contingency table characterising the passengers based on survival and based on the passenger class. (Hint: The R functions <span style="color:blue">`xtabs()`</span> and <span style="color:blue">`addmargins()`</span> may be helpful)

Q2b. Visualize your table using a Bar plot.

Q2c. How many first-class passengers survived the sinking of the Titanic?  (Hint: The R functions <span style="color:blue">`subset()`</span> and <span style="color:blue">`nrow()`</span> may be helpful)

Q2d. What was the percentage of first-class passengers who survived the sinking of the Titanic? (Hint: The R functions <span style="color:blue">`xtabs()`</span> and <span style="color:blue">`prop.table()`</span> may be helpful)




Q2a. Create a two-way contingency table characterising the passengers based on survival and based on the passenger class.
========================================================

```{r}
mytab <- xtabs(~ Survived + Pclass, data=titanic)
# sum of the passengers based on the passengers class and survived
addmargins(mytab)
```

Q2b. Visualize your table using a Bar-plot.
========================================================

```{r, eval=FALSE}
titanic$Pclass <- as.factor(titanic$Pclass)
# frequency counts of the passengers who survived
barplot(mytab, 
        main="Survival by Passenger Class", 
        xlab="Passenger Class", ylab="Frequency",
        col=c("grey", "black"),
        legend=c("Died", "Survived"), beside=TRUE)
```

Q2b. Visualize your table using a Bar-plot.
========================================================

```{r, echo=FALSE, out.height="800px",out.width="800px", fig.align='center'}
titanic$Pclass <- as.factor(titanic$Pclass)
# frequency counts of the passengers who survived
barplot(mytab, 
        main="Survival by Passenger Class", 
        xlab="Passenger Class", ylab="Frequency",
        col=c("grey", "black"),
        legend=c("Died", "Survived"), beside=TRUE)
```


Q2c. How many first-class passengers survived the sinking of the Titanic?
========================================================

```{r}
# using subset() function 
nrow(subset(titanic, Survived == 1 & Pclass == 1))

# using which() function
length(which(Survived == 1 & Pclass == 1))
```

Q2d. What was the percentage of first-class passengers who survived the sinking of the Titanic?
========================================================

```{r}
surviversByClass <- xtabs(~ Survived + Pclass, data=titanic)

# proportion of first-class passengers who survived the sinking of the Titanic  
prop.table(surviversByClass, 2)

# percentage of first-class passengers who survived the sinking of the Titanic  
100*prop.table(surviversByClass, 2)[2,1]
```


QUESTION 3
========================================================

Q3a. Create a three-way contingency table showing the number of passengers based on the passenger's class, gender and survival.  (Hint: The R functions <span style="color:blue">`xtabs()`</span>, <span style="color:blue">`addmargins()`</span> and <span style="color:blue">`ftable()`</span> may be helpful)

Q3b. Express Q3a. in percentages, displaying answers up to two decimal places. (Hint: The R functions <span style="color:blue">`round()`</span>, <span style="color:blue">`prop.table()` and <span style="color:blue">`ftable()`</span> may be helpful)

**Challenge Question C1: **
Visualize your table in Q3b, using a bar plot.

Q3c. How many Females traveling by First-Class survived the sinking of the Titanic?

Q3d. What was the percentage of survivors who were female?
(Hint: Q3c. and Q3d. are not identical)

**Challenge Question C2: **
Visualize your answer in Q3d using a Pie-chart.

Q3e. What was the percentage of females on board the Titanic who survived?

**Challenge Question C3:**
Visualize your answer in Q3e using a Pie-chart.


Q3a. Create a three-way contingency table showing the number of passengers based on the passengers Survived, Sex and Passenger Class.
========================================================

```{r, eval=FALSE}
# three way contingency table
# Survived, Sex and Passenger Class
mytable1 <- xtabs(~ Survived + Sex + Pclass, data=titanic)
addmargins(mytable1)
```

Q3a. Create a three-way contingency table showing the number of passengers based on the passengers Survived, Sex and Passenger Class.
========================================================

```{r, echo=FALSE}
# three way contingency table
# Survived, Sex and Passenger Class
mytable1 <- xtabs(~ Survived + Sex + Pclass, data=titanic)
addmargins(mytable1)
```

Q3a. Create a three-way contingency table showing the number of passengers based on the Survived, Sex and Passenger Class.
========================================================

```{r}
# frquency counts
ftable(mytable1)
```

Q3a. Express the previous question in percentages.
========================================================

```{r}
# frequency counts of the passengers
ftab <- ftable(prop.table(mytable1))

# converting to percentages
ftabPer <- ftab*100

# rounding values to 2 decimal places.
round(ftabPer, 2)
```


Q3a. Create a three-way contingency table showing the number of passengers based on the Survival, Passenger Class and Sex.
========================================================

```{r, eval=FALSE}
# three way contingency table
# Survived, Passenger Classs and Sex
mytable2 <- xtabs(~ Survived + Pclass + Sex, data=titanic)
addmargins(mytable2)
```

Q3a. Create a three-way contingency table showing the number of passengers based on the Survival, Passenger Class and Sex.
========================================================

```{r, echo=FALSE}
# three way contingency table
# Survived, Passenger Classs and Sex
mytable2 <- xtabs(~ Survived + Pclass + Sex, data=titanic)
addmargins(mytable2)
```

Q3a. Create a three-way contingency table showing the number of passengers based on the Survival, Passenger Class and Sex.
========================================================

```{r}
# frquency counts
ftable(mytable2)
```

Q3b. Express the previous question in percentages.
========================================================

```{r}
# frequency counts of the passengers
ftab <- ftable(prop.table(mytable2))

# converting to percentages
ftabPer <- ftab*100

# rounding values to 2 decimal places.
round(ftabPer, 2)
```


Challenge Question C1: Visualize your table using Barplot.
========================================================

```{r, eval=FALSE, out.height="650px",out.width="700px", fig.align='center'}
# percent of total of the passengers who survived
par(mfrow=c(1,2),mar=c(4,4,3,3))
tab2 <- xtabs(~ Survived + Pclass + Sex, data=titanic)
barplot(tab2[,,1],
	main=dimnames(tab2)$Sex[1],
	beside=TRUE,
	ylim=c(0,400),
	col=c("grey","black"),
	ylab="Number of Passengers",
	legend=c("Died", "Survived"),
	args.legend=list(x="topleft"))

barplot(tab2[,,2],
	main=dimnames(tab2)$Sex[2],
	beside=TRUE,
	ylim=c(0,400),
	col=c("grey","black"),
	#ylab="Number of Passengers",
	legend=c("Died", "Survived"),
	args.legend=list(x="topleft"))
```

Challenge Question C1: Visualize your table using Barplot.
========================================================

```{r, echo=FALSE, out.height="650px",out.width="800px", fig.align='center'}
# percent of total of the passengers who survived
par(mfrow=c(1,2),mar=c(4,4,3,3))
tab2 <- xtabs(~ Survived + Pclass + Sex, data=titanic)
barplot(tab2[,,1],
	main=dimnames(tab2)$Sex[1],
	beside=TRUE,
	ylim=c(0,400),
	col=c("grey","black"),
	ylab="Number of Passengers",
	legend=c("Died", "Survived"),
	args.legend=list(x="topleft"), las=0)
  
barplot(tab2[,,2],
	main=dimnames(tab2)$Sex[2],
	beside=TRUE,
	ylim=c(0,400),
	col=c("grey","black"),
	#ylab="Number of Passengers",
	legend=c("Died", "Survived"),
	args.legend=list(x="topleft"), las=0)
```

Challenge Question C1: Visualize your table using Barplot.
========================================================

```{r, eval=FALSE}
# percent of total of the passengers who survived
par(mfrow=c(1,3),mar=c(8,4,3,3))
tab2 <- xtabs(~ Survived + Sex + Pclass, data=titanic)
barplot(tab2[,,1],
	main=dimnames(tab2)$Pclass[1],
	beside=TRUE,
	ylim=c(0,350),
	col=c("grey","black"),
	ylab="Number of Passengers",
	legend=c("Died", "Survived"),
	args.legend=list(x="topleft"))

barplot(tab2[,,2],
	main=dimnames(tab2)$Pclass[2],
	beside=TRUE,
	ylim=c(0,350),
	col=c("grey","black"),
	#ylab="Number of Passengers",
	legend=c("Died", "Survived"),
	args.legend=list(x="topleft"))

barplot(tab2[,,3],
	main=dimnames(tab2)$Pclass[3],
	beside=TRUE,
	ylim=c(0,350),
	col=c("grey","black"),
	#ylab="Number of Passengers",
	legend=c("Died", "Survived"),
	args.legend=list(x="topleft"))
```

Challenge Question C1: Visualize your table using Barplot.
========================================================

```{r, echo=FALSE, out.height="800px",out.width="1000px", fig.align='center'}
# percent of total of the passengers who survived
par(mfrow=c(1,3),mar=c(8,4,3,3))
tab2 <- xtabs(~ Survived + Sex + Pclass, data=titanic)
barplot(tab2[,,1],
	main=dimnames(tab2)$Pclass[1],
	beside=TRUE,
	ylim=c(0,350),
	col=c("grey","black"),
	ylab="Number of Passengers",
	legend=c("Died", "Survived"),
	args.legend=list(x="topleft"))

barplot(tab2[,,2],
	main=dimnames(tab2)$Pclass[2],
	beside=TRUE,
	ylim=c(0,350),
	col=c("grey","black"),
	#ylab="Number of Passengers",
	legend=c("Died", "Survived"),
	args.legend=list(x="topleft"))

barplot(tab2[,,3],
	main=dimnames(tab2)$Pclass[3],
	beside=TRUE,
	ylim=c(0,350),
	col=c("grey","black"),
	#ylab="Number of Passengers",
	legend=c("Died", "Survived"),
	args.legend=list(x="topleft"))
```


Q3c. How many Females traveling by First-Class survived the sinking of the Titanic?
========================================================

```{r}
# females traveling by first-class survived the sinking of the Titanic
ftable(mytable1)[3]
```

Q3d. What was the percentage of survivors who were female?
========================================================

```{r}
surviversBySex <- xtabs(~ Survived + Sex, data=titanic)
# frequency counts of survivors who were female
surviversBySex

# proportions
propSur <- prop.table(surviversBySex,1)

# percentage
propSurPer <- propSur*100

# percentage of survivors who were female
propSurPer[2,1]
```

Challenge Question C2: Visualize your answer in Q3d using a Pie-chart.
========================================================

```{r, eval=FALSE}
slices <- c(67.94118, 32.05882)
lbls <- c("Female survivers", "Female who died")

# percentage of female who survived and who died
pct <- round(slices/sum(slices)*100)                      
lbls <- paste(lbls, pct) 
lbls <- paste(lbls,"%",sep="")
pie(slices,labels = lbls, col = c("grey", "black"),
    main="Pie Chart with Percentages")
```

Challenge Question C2: Visualize your answer in Q3d using a Pie-chart.
========================================================

```{r, echo=FALSE, out.height="650px",out.width="700px", fig.align='center'}
slices <- c(67.94118, 32.05882)
lbls <- c("Female survivers", "Female who died")

# percentage of female who survived and who died
pct <- round(slices/sum(slices)*100)                      
lbls <- paste(lbls, pct) 
lbls <- paste(lbls,"%",sep="")
pie(slices,labels = lbls, col = c("grey", "black"),
    main="Pie Chart with Percentages")
```

Q3e. What was the percentage of females on board the Titanic who survived?
========================================================

```{r}
# proportion
propSur2 <- prop.table(surviversBySex,2)

# percentage
propSur2Per <- propSur2*100

# percentage of total females on the Titanic who survived
propSur2Per
```

Challenge Question C3: Visualize your answer in Q3f using a Pie-chart.
========================================================

```{r, eval=FALSE}
slices <- c(74.03846, 25.96154)
lbls <- c("Total female survivers", "Total female who died")

# percentage of total female who survived and who died
pct <- round(slices/sum(slices)*100)                      
lbls <- paste(lbls, pct) 
lbls <- paste(lbls,"%",sep="")
pie(slices,labels = lbls, col = c("grey", "black"),
    main="Pie Chart with Percentages")
```

Challenge Question C3: Visualize your answer in Q3f using a Pie-chart.
========================================================

```{r, echo=FALSE, out.height="650px",out.width="700px", fig.align='center'}
slices <- c(74.03846, 25.96154)
lbls <- c("Total female survivers", "Total female who died")

# percentage of total female who survived and who died
pct <- round(slices/sum(slices)*100)                      
lbls <- paste(lbls, pct) 
lbls <- paste(lbls,"%",sep="")
pie(slices,labels = lbls, col = c("grey", "black"),
    main="Pie Chart with Percentages")
```


QUESTION 4
========================================================

Q4a. Use a Pearson's Chi-squared test to evaluate whether the proportion of females who survived was larger than the proportion of males who survived?

Q4b. What is the p-value of the previous Pearson's Chi-squared test?

**Challenge Question C4: **
Learn about Mosaic Plot by browsing this link. 
Create a Mosaic Plot of Titanic survivors and nonsurvivors based on gender (male/female), passenger class (First/Second/Third).


Q4a. Use a Pearson's Chi-squared test to evaluate whether the proportion of females who survived was larger than the proportion of males who survived?
========================================================

```{r}
chisq.test(surviversBySex)
```

Percentage of female survivers was higher than percentage of male survivers.

Q4b. What is the p-value of the previous Pearson's Chi-squared test?
========================================================

```{r}
# p-value
(chisq.test(surviversBySex))$p.value
```
The p-value of the previous Pearson's Chi-squared test is $3.77991e-58$.

Challenge Question C4: Create a Mosaic Plot of Titanic survivors and nonsurvivors based on gender (male/female), passenger class (First/Second/Third).
========================================================

Create a Mosaic Plot of Titanic survivors and nonsurvivors based on gender (male/female), passenger class (First/Second/Third).

```{r, eval=FALSE}
library(vcd)
# Passenger Class, Sex and Survived
mosaic(~ Pclass + Sex + Survived, data=titanic, shade=TRUE, legend=TRUE)
```

Challenge Question C4: Create a Mosaic Plot of Titanic survivors and nonsurvivors based on gender (male/female), passenger class (First/Second/Third).
========================================================

```{r, echo=FALSE, out.height="800px",out.width="800px", fig.align='center'}
library(vcd)
# Passenger Class, Sex and Survived
mosaic(~ Pclass + Sex + Survived, data=titanic, shade=TRUE, legend=TRUE)
```


Challenge Question C4: Create a Mosaic Plot of Titanic survivors and nonsurvivors based on gender (male/female), passenger class (First/Second/Third).
========================================================

Create a Mosaic Plot of Titanic survivors and nonsurvivors based on gender (male/female), passenger class (First/Second/Third).

```{r, eval=FALSE}
library(vcd)
# Sex, Passenger Class and Survived
mosaic(~ Sex + Pclass + Survived, data=titanic, shade=TRUE, legend=TRUE)
```

Challenge Question C4: Create a Mosaic Plot of Titanic survivors and nonsurvivors based on gender (male/female), passenger class (First/Second/Third).
========================================================

Create a Mosaic Plot of Titanic survivors and nonsurvivors based on gender (male/female), passenger class (First/Second/Third).

```{r, echo=FALSE, echo=FALSE, out.height="700px",out.width="700px", fig.align='center'}
library(vcd)
# Sex, Passenger Class and Survived
mosaic(~ Sex + Pclass + Survived, data=titanic, shade=TRUE, legend=TRUE)
```


Challenge Question C4: Create a Mosaic Plot of Titanic survivors and nonsurvivors based on gender (male/female), passenger class (First/Second/Third).
========================================================

Create a Mosaic Plot of Titanic survivors and nonsurvivors based on gender (male/female), passenger class (First/Second/Third).

```{r, eval=FALSE}
library(vcd)
# Survived, Passenger Class and Sex
mosaic(~ Survived + Pclass + Sex, data=titanic, shade=TRUE, legend=TRUE)
```

Challenge Question C4: Create a Mosaic Plot of Titanic survivors and nonsurvivors based on gender (male/female), passenger class (First/Second/Third).
========================================================

Create a Mosaic Plot of Titanic survivors and nonsurvivors based on gender (male/female), passenger class (First/Second/Third).

```{r, echo=FALSE, echo=FALSE, out.height="700px",out.width="700px", fig.align='center'}
library(vcd)
# Survived, Passenger Class and Sex
mosaic(~ Survived + Pclass + Sex, data=titanic, shade=TRUE, legend=TRUE)
```


